---
title: Approximate MNN correction with interchangeable BNPARAM backends
author:
- name: Oz Beker
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
package: batchelor
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{3. Approximate MNN correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE, autodep=TRUE)
```

```{r setup, echo=FALSE, message=FALSE}
library(batchelor)
library(BiocNeighbors)
library(scRNAseq)
library(scuttle)
library(scran)
library(ggplot2)
```

# Introduction

The `fastMNN()` and `mnnCorrect()` functions in `r Biocpkg("batchelor")`
delegate nearest-neighbor search to `BiocNeighbors` through the `BNPARAM`
argument. This makes it possible to compare exact and approximate
nearest-neighbor backends inside the same MNN correction framework, rather than
maintaining a separate ANN-specific correction function.

This vignette is written to be extensible. All method-specific work is driven by
a method registry near the top of the document, so future backends can be added
by appending one new entry rather than rewriting the benchmarking or plotting
code. For now, we compare:

- `KmknnParam()` as the exact baseline.
- `HnswParam()` as a graph-based approximate method.
- `NndescentParam()` as an ANN-descent approximate method.
- `AnnoyParam()` as a forest-based approximate method.
- `RpforestParam()` as a random projection forest approximate method.

This version of the vignette is set up for fast smoke testing. It focuses on
the `fastMNN()` workflow, avoids the much slower classic `mnnCorrect()` path,
and runs a single benchmark seed by default. The helper code is still written
so that we can later expand back to a 5-seed summary without changing the rest
of the notebook.

All standalone figure PDFs are written to a sidecar directory so they can be
edited later.

```{r helpers, echo=FALSE}
full_benchmark_seeds <- c(101L, 202L, 303L, 404L, 505L)
benchmark_seeds <- full_benchmark_seeds[1]

method_registry <- data.frame(
    id=c("kmknn", "hnsw", "nndescent", "annoy", "rpforest"),
    label=c("Kmknn (exact)", "HNSW", "NN-descent", "Annoy", "RP forest"),
    reference=c(TRUE, FALSE, FALSE, FALSE, FALSE),
    stringsAsFactors=FALSE
)
method_registry$param <- I(list(
    function() KmknnParam(),
    function() HnswParam(),
    function() NndescentParam(),
    function() AnnoyParam(),
    function() RpforestParam()
))
backend_files <- file.path(
    "BiocNeighbors", "R",
    c("KmknnParam.R", "HnswParam.R", "NndescentParam.R", "AnnoyParam.R", "RpforestParam.R")
)
backend_stamp <- paste(file.info(backend_files[file.exists(backend_files)])$mtime, collapse="|")
method_cache_token <- paste(method_registry$id, method_registry$label, backend_stamp, collapse="|")

reference_id <- method_registry$id[method_registry$reference]
stopifnot(length(reference_id) == 1L)

instantiate_param <- function(method_id) {
    method_registry$param[[match(method_id, method_registry$id)]]()
}

method_factor <- function(x) {
    factor(x, levels=method_registry$id, labels=method_registry$label)
}

non_reference_methods <- function() {
    method_registry$id[!method_registry$reference]
}

resolve_figure_dir <- function() {
    input <- tryCatch(knitr::current_input(dir=TRUE), error=function(e) "")
    if (nzchar(input)) {
        repo_root <- dirname(dirname(normalizePath(input)))
    } else {
        repo_root <- normalizePath(getwd())
    }
    out <- Sys.getenv(
        "BATCHELOR_APPROX_MNN_FIG_DIR",
        unset=file.path(repo_root, "vignettes", "figures", "approx-mnn")
    )
    dir.create(out, recursive=TRUE, showWarnings=FALSE)
    out
}

figure_dir <- resolve_figure_dir()

point_geom <- function(...) {
    if (requireNamespace("ggrastr", quietly=TRUE)) {
        ggrastr::geom_point_rast(..., dpi=300, dev="cairo")
    } else {
        ggplot2::geom_point(...)
    }
}

save_sidecar_pdf <- function(plot, filename, width=8, height=5) {
    ggplot2::ggsave(
        filename=file.path(figure_dir, filename),
        plot=plot,
        device=grDevices::cairo_pdf,
        width=width,
        height=height,
        units="in",
        dpi=300,
        bg="white"
    )
    invisible(plot)
}

benchmark_methods <- function(method_ids, seeds, prepare_once, run_once) {
    collected <- vector("list", length(method_ids) * length(seeds))
    counter <- 1L

    for (seed in seeds) {
        prepared <- prepare_once(seed)
        for (method in method_ids) {
            set.seed(seed)
            elapsed <- unname(system.time(
                run_once(prepared, instantiate_param(method))
            )[["elapsed"]])
            collected[[counter]] <- data.frame(
                seed=seed,
                method=method,
                seconds=elapsed,
                stringsAsFactors=FALSE
            )
            counter <- counter + 1L
        }
        rm(prepared)
        invisible(gc())
    }

    do.call(rbind, collected)
}

summarize_timings <- function(x, by) {
    key <- x[, by, drop=FALSE]
    mean.df <- aggregate(x$seconds, key, mean)
    sd.df <- aggregate(x$seconds, key, stats::sd)
    names(mean.df)[ncol(mean.df)] <- "mean.seconds"
    names(sd.df)[ncol(sd.df)] <- "sd.seconds"
    output <- merge(mean.df, sd.df, by=by, sort=FALSE)
    output$sd.seconds[is.na(output$sd.seconds)] <- 0
    output
}

make_regime_label <- function(space, dims, unit) {
    paste0(space, " (", dims, " ", unit, if (dims == 1L) "" else "s", ")")
}

recall_per_cell <- function(approx, exact) {
    vapply(seq_len(nrow(approx)), function(i) {
        length(intersect(approx[i, ], exact[i, ])) / ncol(exact)
    }, numeric(1))
}

jaccard_per_cell <- function(approx, exact) {
    vapply(seq_len(nrow(approx)), function(i) {
        a <- approx[i, ]
        e <- exact[i, ]
        length(intersect(a, e)) / length(union(a, e))
    }, numeric(1))
}

compute_overlap_metrics <- function(data, regime_label, k.fixed=20L, k.grid=c(5L, 10L, 20L, 40L, 80L)) {
    set.seed(999)
    ref.fixed <- findKNN(data, k=k.fixed, BNPARAM=instantiate_param(reference_id), get.distance=FALSE)$index

    dist.list <- lapply(non_reference_methods(), function(method) {
        set.seed(999)
        approx.fixed <- findKNN(data, k=k.fixed, BNPARAM=instantiate_param(method), get.distance=FALSE)$index

        rbind(
            data.frame(
                regime=regime_label,
                method=method,
                metric="Recall@k",
                value=recall_per_cell(approx.fixed, ref.fixed),
                stringsAsFactors=FALSE
            ),
            data.frame(
                regime=regime_label,
                method=method,
                metric="Jaccard@k",
                value=jaccard_per_cell(approx.fixed, ref.fixed),
                stringsAsFactors=FALSE
            )
        )
    })
    dist.df <- do.call(rbind, dist.list)

    k.list <- lapply(k.grid, function(kk) {
        set.seed(999)
        ref.current <- findKNN(data, k=kk, BNPARAM=instantiate_param(reference_id), get.distance=FALSE)$index

        do.call(rbind, lapply(non_reference_methods(), function(method) {
            set.seed(999)
            approx.current <- findKNN(data, k=kk, BNPARAM=instantiate_param(method), get.distance=FALSE)$index
            data.frame(
                regime=regime_label,
                method=method,
                k=kk,
                metric=c("Recall@k", "Jaccard@k"),
                mean=c(
                    mean(recall_per_cell(approx.current, ref.current)),
                    mean(jaccard_per_cell(approx.current, ref.current))
                ),
                stringsAsFactors=FALSE
            )
        }))
    })

    list(
        distribution=dist.df,
        by_k=do.call(rbind, k.list)
    )
}

make_synth_gene_batches <- function(total_cells, ngenes=200L, shift=0.75, seed=1) {
    set.seed(seed)
    n_per_batch <- total_cells %/% 2L
    centers <- rbind(
        seq(-2, 2, length.out=ngenes),
        c(rep(-1.25, ngenes %/% 2L), rep(1.25, ngenes - ngenes %/% 2L)),
        c(rep(1.5, ngenes %/% 4L), rep(0, ngenes %/% 2L), rep(-1.5, ngenes - (ngenes %/% 4L + ngenes %/% 2L)))
    )
    batch_shift <- seq(-shift, shift, length.out=ngenes)

    clust1 <- sample(seq_len(nrow(centers)), n_per_batch, replace=TRUE)
    clust2 <- sample(seq_len(nrow(centers)), n_per_batch, replace=TRUE)

    B1 <- centers[clust1, , drop=FALSE] +
        matrix(rnorm(n_per_batch * ngenes, sd=0.6), n_per_batch, ngenes)
    B2 <- centers[clust2, , drop=FALSE] +
        matrix(rep(batch_shift, each=n_per_batch), n_per_batch, ngenes) +
        matrix(rnorm(n_per_batch * ngenes, sd=0.6), n_per_batch, ngenes)

    list(B1=t(B1), B2=t(B2))
}
```

The current sidecar figure directory is:

```{r figure-dir}
normalizePath(figure_dir, winslash="/", mustWork=FALSE)
```

# Load, QC, normalization and HVG selection

We follow the same preprocessing pattern as the main
[correction vignette](correction.html), using the Zeisel and Tasic brain data
sets from `r Biocpkg("scRNAseq")`.

```{r load-data}
sce1 <- ZeiselBrainData()
sce2 <- TasicBrainData()
```

We apply quick quality control to both data sets.

```{r qc}
sce1 <- addPerCellQC(sce1, subsets=list(Mito=grep("mt-", rownames(sce1))))
qc1 <- quickPerCellQC(colData(sce1), sub.fields="subsets_Mito_percent")
sce1 <- sce1[, !qc1$discard]

sce2 <- addPerCellQC(sce2, subsets=list(Mito=grep("mt_", rownames(sce2))))
qc2 <- quickPerCellQC(colData(sce2), sub.fields="subsets_Mito_percent")
sce2 <- sce2[, !qc2$discard]
```

We restrict to the common gene universe and perform multi-batch normalization.

```{r norm}
universe <- intersect(rownames(sce1), rownames(sce2))
sce1 <- sce1[universe, ]
sce2 <- sce2[universe, ]

normed <- multiBatchNorm(sce1, sce2)
sce1 <- normed[[1]]
sce2 <- normed[[2]]
```

Finally, we choose the top 5000 highly variable genes.

```{r hvgs}
dec1 <- modelGeneVar(sce1)
dec2 <- modelGeneVar(sce2)
combined.dec <- combineVar(dec1, dec2)
chosen.hvgs <- getTopHVGs(combined.dec, n=5000)
reduced_dims <- 50L
gene_dims_example <- length(chosen.hvgs)

data.frame(
    batch=c("Zeisel", "Tasic"),
    cells=c(ncol(sce1), ncol(sce2))
)
```

For neighbor-set diagnostics later on, we also precompute a shared
cosine-normalized PCA space.

```{r prepare-pcs, cache=TRUE, cache.extra=method_cache_token}
cos1 <- cosineNorm(SummarizedExperiment::assay(sce1, "logcounts"))
cos2 <- cosineNorm(SummarizedExperiment::assay(sce2, "logcounts"))
pcs <- multiBatchPCA(
    cos1, cos2,
    subset.row=chosen.hvgs,
    d=reduced_dims,
    BSPARAM=BiocSingular::ExactParam()
)
```

# Timing the methods

We benchmark `fastMNN()` across all backends in the registry. The current smoke
test uses a single seed via `benchmark_seeds`; later, this can be switched back
to `full_benchmark_seeds` to recover 5-seed mean +/- SD summaries.

```{r example-benchmarks, cache=TRUE, cache.extra=method_cache_token}
example_timings <- benchmark_methods(
    method_ids=method_registry$id,
    seeds=benchmark_seeds,
    prepare_once=function(seed) list(B1=sce1, B2=sce2),
    run_once=function(prepared, param) {
        invisible(fastMNN(prepared$B1, prepared$B2, subset.row=chosen.hvgs, BNPARAM=param))
    }
)
example_summary <- summarize_timings(example_timings, by="method")
example_summary$method <- method_factor(example_summary$method)
example_summary
```

```{r example-timing-plot, fig.width=8, fig.height=4.8, fig.cap="Runtime of exact and approximate backends on the Zeisel and Tasic example data using fastMNN(). Bars show the current smoke-test summary; with one seed the SD bars collapse to zero."}
timing_plot <- ggplot(example_summary, aes(method, mean.seconds, fill=method)) +
    geom_col(width=0.7) +
    geom_errorbar(
        aes(ymin=pmax(mean.seconds - sd.seconds, 0), ymax=mean.seconds + sd.seconds),
        width=0.15
    ) +
    labs(
        x=NULL,
        y="elapsed time (s)",
        subtitle=paste0("Input: ", length(chosen.hvgs), " HVGs, PCA dimension = ", reduced_dims,
            ", benchmark seeds = ", paste(benchmark_seeds, collapse=", "))
    ) +
    theme_minimal() +
    theme(
        legend.position="none",
        axis.text.x=element_text(angle=25, hjust=1)
    )

save_sidecar_pdf(timing_plot, "01-example-timing.pdf", width=8, height=4.8)
timing_plot
```

# Scaling to more cells

To compare scaling behavior, we generate synthetic matched batches at a range of
total cell counts and rerun `fastMNN()` with each backend. We keep the
synthetic gene-space dimensionality at 200 and use `d = 50` for the fastMNN PCA
step so the benchmark remains practical to render.

```{r scaling-grid}
sizes <- c(1000L, 3000L, 10000L, 30000L, 100000L)
synthetic_gene_dims <- 200L
```

```{r scaling-benchmarks, cache=TRUE, cache.extra=method_cache_token}
scaling_timings <- do.call(rbind, lapply(sizes, function(total_cells) {
    current <- benchmark_methods(
        method_ids=method_registry$id,
        seeds=benchmark_seeds,
        prepare_once=function(seed) make_synth_gene_batches(total_cells, ngenes=synthetic_gene_dims, seed=seed),
        run_once=function(prepared, param) {
            invisible(fastMNN(prepared$B1, prepared$B2, d=reduced_dims, cos.norm=FALSE, BNPARAM=param))
        }
    )
    current$total_cells <- total_cells
    current
}))
scaling_summary <- summarize_timings(scaling_timings, by=c("method", "total_cells"))
scaling_summary$method <- method_factor(scaling_summary$method)
scaling_summary
```

```{r scaling-plot, fig.width=8, fig.height=5, fig.cap="Scaling of exact and approximate backends with increasing total cell count using fastMNN() on synthetic batches. Lines show the current smoke-test summary; with one seed the SD bars collapse to zero."}
scaling_plot <- ggplot(
    scaling_summary,
    aes(total_cells, mean.seconds, colour=method, group=method)
) +
    geom_line(linewidth=0.8) +
    point_geom(size=2.2) +
    geom_errorbar(
        aes(ymin=pmax(mean.seconds - sd.seconds, 0), ymax=mean.seconds + sd.seconds),
        width=0
    ) +
    labs(
        x="total cells",
        y="elapsed time (s)",
        colour="backend",
        subtitle=paste0("Synthetic input: ", synthetic_gene_dims, " genes, fastMNN d = ", reduced_dims,
            ", benchmark seeds = ", paste(benchmark_seeds, collapse=", "))
    ) +
    theme_minimal()

save_sidecar_pdf(scaling_plot, "02-scaling.pdf", width=8, height=5)
scaling_plot
```

# Approximate versus exact neighbor set comparison

We now compare the exact and approximate neighbor sets directly in the shared
PCA space used by the fast workflow. The exact reference is the backend marked
as `reference=TRUE` in the method registry, which is currently `KmknnParam()`.
Each approximate backend is compared against that exact reference with two
set-overlap metrics:

- `Recall@k`, the fraction of exact neighbors recovered by the approximate set.
- `Jaccard@k`, the overlap between the exact and approximate sets divided by
  the size of their union.

We compute these metrics in the shared reduced space on the same preprocessed
example data.

```{r neighbor-spaces, cache=TRUE, cache.extra=method_cache_token}
reduced_joint <- rbind(pcs[[1]], pcs[[2]])

reduced_overlap <- compute_overlap_metrics(
    reduced_joint,
    regime_label=make_regime_label("Reduced space", reduced_dims, "dimension")
)

overlap_distribution <- reduced_overlap$distribution
overlap_by_k <- reduced_overlap$by_k

overlap_distribution$method <- method_factor(overlap_distribution$method)
overlap_distribution$regime <- factor(
    overlap_distribution$regime,
    levels=make_regime_label("Reduced space", reduced_dims, "dimension")
)
overlap_distribution$metric <- factor(overlap_distribution$metric, levels=c("Recall@k", "Jaccard@k"))

overlap_by_k$method <- method_factor(overlap_by_k$method)
overlap_by_k$regime <- factor(
    overlap_by_k$regime,
    levels=levels(overlap_distribution$regime)
)
overlap_by_k$metric <- factor(overlap_by_k$metric, levels=levels(overlap_distribution$metric))
```

## Per-cell distributions

```{r overlap-distribution-plot, fig.width=8.5, fig.height=4.8, fig.cap="Per-cell recall and Jaccard overlap of approximate neighbor sets against the exact Kmknn baseline at k = 20 in the shared PCA space."}
distribution_plot <- ggplot(
    overlap_distribution,
    aes(method, value, fill=method)
) +
    geom_violin(scale="width", alpha=0.6) +
    geom_boxplot(width=0.15, outlier.size=0.3, alpha=0.8) +
    facet_grid(regime ~ metric) +
    coord_cartesian(ylim=c(0, 1)) +
    labs(x=NULL, y="per-cell overlap") +
    theme_minimal() +
    theme(
        legend.position="none",
        axis.text.x=element_text(angle=25, hjust=1)
    )

save_sidecar_pdf(distribution_plot, "03-overlap-distribution.pdf", width=8.5, height=4.8)
distribution_plot
```

## Empirical cumulative distributions

```{r overlap-ecdf-plot, fig.width=8.5, fig.height=4.8, fig.cap="Empirical cumulative distributions of per-cell recall and Jaccard overlap for the approximate backends in the shared PCA space."}
ecdf_plot <- ggplot(
    overlap_distribution,
    aes(value, colour=method)
) +
    stat_ecdf(geom="step", linewidth=0.9) +
    facet_grid(regime ~ metric) +
    coord_cartesian(xlim=c(0, 1), ylim=c(0, 1)) +
    labs(x="per-cell overlap", y="cumulative fraction of cells", colour="backend") +
    theme_minimal()

save_sidecar_pdf(ecdf_plot, "04-overlap-ecdf.pdf", width=8.5, height=4.8)
ecdf_plot
```

## Mean overlap as a function of k

```{r overlap-by-k-plot, fig.width=8.5, fig.height=4.8, fig.cap="Mean recall and Jaccard overlap across a grid of neighborhood sizes in the shared PCA space."}
overlap_k_plot <- ggplot(
    overlap_by_k,
    aes(k, mean, colour=method, group=method)
) +
    geom_line(linewidth=0.8) +
    point_geom(size=2.1) +
    facet_grid(regime ~ metric) +
    coord_cartesian(ylim=c(0, 1)) +
    labs(x="k", y="mean overlap", colour="backend") +
    theme_minimal()

save_sidecar_pdf(overlap_k_plot, "05-overlap-by-k.pdf", width=8.5, height=4.8)
overlap_k_plot
```

# Session information

```{r session}
sessionInfo()
```

# References {-}
